Abstract Binding Trees

### 2015-10-25 paper.tex
Split f into two explicit maps
Begin implementing case study (#4)

\newcommand\MkValence[3]{\{#1\}[#2].\, #3}
\newcommand\MkBTm[3]{\{#1\}[#2].\, #3}
\newcommand\MkArity[2]{(#1) #2}
\newcommand\IsArity[1]{\IMode{#1}\ \textit{arity}}
\newcommand\IsValence[1]{\IMode{#1}\ \textit{valence}}
\newcommand\IsSort[1]{\IMode{#1}\ \textit{sort}}
\newcommand\Leq[2]{\IMode{#1}\leq\IMode{#2}}

@@

\newcommand\Grothendieck[2]{\oint^{#1} #2#1}
\newcommand\Pair[2]{\langle #1, #2 \rangle}
\newcommand\Lift[2]{
  \Operators\langle#1\ifthenelse{\equal{#2}{}}{}{, #2}\rangle
}
\newcommand\PairTimes[2]{\langle #1\, {\scriptstyle {{\times}}}\, #2 \rangle}
\newcommand\Singleton{\bullet}
\newcommand\Forget[1]{|\, #1\, |}

@@

Then, fix a covariant presheaf (copresheaf) of operators
$\Of{\Operators}{\SCtx\times\Discrete{\Arities}\to\Sets}$ such that the arrows
in $\SCtx$ lift to renamings of operators' parameters. Via the Grothendieck
construction\footnote{In this case, $C\Grothendieck{}{\Psi}$ represents the
  category of elements of a copresheaf $\Of{\Psi}{C\to\Sets}$ but we keep the
  $C$ implicit and simply refer to it as the Grothendieck construction.
  Alternatively, this construction can be understood as a coend
$\IMode{C\Grothendieck{}{\Psi}}\cong \IMode{\int^{c \in C}\Coslice{c}{C}
\otimes \Discrete{\Psi(c)}}$.} $\Of{\Grothendieck{}{(-)}}{\Sets^{C}\to\Cats}$
on operators, we have a category of objects
$\Member{\Pair{\Pair{\Upsilon}{a}}{\vartheta}}{\Grothendieck{}{\Operators}}$
for $\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$ and morphisms
$\Hom{\Grothendieck{}{\Operators}}{\Pair{\Pair{\Upsilon}{a}}{\vartheta}}{\Pair{\Pair{\Upsilon'}{a'}}{\vartheta'}}$
for $\Of{f}{\Pair{\Upsilon}{a}\to\Pair{\Upsilon'}{a'}}$ such that
$\EqMember{\Lift{f}{}(\vartheta)}{\vartheta'}{\Operators\Pair{\Upsilon'}{a'}}$.
Equivalently, $\Grothendieck{}{\Operators}$ is the pullback of $\Operators$
along the universal $\Sets$-bundle where $\pi_{\Operators}$ is a discrete
Grothendieck opfibration and $\Forget{-}_{\Singleton}$ is the forgetful functor
from pointed sets:

@@

Operators are defined by specifying the fibers of $\pi_{\Operators}$ in which
they reside. For instance, consider the lambda calculus with a single sort,
$\SortExp$; we give its signature by asserting the following about its
operators:
\begin{gather*}
  \IsOperator{\Upsilon}{\lambda}{
    \MkArity{\MkValence{\cdot}{\SortExp}{\SortExp}}{\SortExp}
  }\\
  \IsOperator{\Upsilon}{\mathsf{ap}}{
    \MkArity{
      \MkValence{\cdot}{\cdot}{\SortExp},
      \MkValence{\cdot}{\cdot}{\SortExp}
    }{\SortExp}
  }
\end{gather*}

@@

\begin{gather*}
  \IsOperator{\Upsilon}{\mathsf{decl}}{
    \MkArity{
      \MkValence{\cdot}{\cdot}{\SortExp},
      \MkValence{\SortExp}{\cdot}{\SortExp}
    }{
      \SortExp
    }
  }\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathsf{get}[u]}{
    \MkArity{
      \cdot
    }{
      \SortExp
    }
  }\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathsf{set}[u]}{
    \MkArity{
      \MkValence{\cdot}{\cdot}{\SortExp}
    }{
      \SortExp
    }
  }
\end{gather*}

@@

Let the judgment $\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{s}$ presuppose
$\IsMetaCtx{\Omega}$, $\IsSymCtx\Upsilon$, $\IsVarCtx\Gamma$ and $\IsSort{s}$,
meaning that $M$ is an abstract binding tree of sort $s$, with metavariables in
$\Omega$, parameters in $\Upsilon$, and variables in $\Gamma$. Let the judgment
$\IsBTm{\Omega}{\Upsilon}{\Gamma}{E}{v}$ presuppose $\IsValence{v}$.  Then, the
syntax of abstract binding trees is inductively defined in four rules:

@@

Abstract binding trees are identified up to $\alpha$-equivalence. Let $\FS{M}$
be the collection of symbols free in $M$, and let $\FV{M}$ be the collection of
variables free $\pi_\Operators$ to operators in $M$. %% TODO: actually define these

@@

Variable substitution in abstract binding trees is defined inductively by a
pair of judgments, $\IsSubst{N}{x}{M}{M'}$ and $\IsSubst{N}{x}{E}{F}$:

@@

>>>>

\newcommand\MkValence[3]{%
  \ifthenelse{\equal{#1}{}}{}{%
    \{#1\}%
  }%
  \ifthenelse{\equal{#2}{}}{}{%
    [#2]%
  }.\, #3
}
\newcommand\MkBTm[3]{\{#1\}[#2].\, #3}
\newcommand\MkArity[2]{(#1) #2}
\newcommand\IsArity[1]{\IMode{#1}\ \textit{arity}}
\newcommand\IsValence[1]{\IMode{#1}\ \textit{valence}}
\newcommand\IsSort[1]{\IMode{#1}\ \textit{sort}}
\newcommand\Leq[2]{\IMode{#1}\leq\IMode{#2}}

@@

\newcommand\Grothendieck[2]{\oint^{#1} #2#1}
\newcommand\Pair[2]{\langle #1, #2 \rangle}
\newcommand\Lift[2]{
  \Operators\langle#1%
  \ifthenelse{\equal{#2}{}}{}{, #2}\rangle
}
\newcommand\PairTimes[2]{\langle #1\, {\scriptstyle {{\times}}}\, #2 \rangle}
\newcommand\Singleton{\bullet}
\newcommand\Forget[1]{|\, #1\, |}

@@

Then, fix a covariant presheaf (copresheaf) of operators
$\Of{\Operators}{\SCtx\times\Discrete{\Arities}\to\Sets}$ such that the arrows
in $\SCtx$ lift to renamings of operators' parameters. Via the Grothendieck
construction\footnote{In this case, $C\Grothendieck{}{\Psi}$ represents the
  category of elements of a copresheaf $\Of{\Psi}{C\to\Sets}$ but we keep the
  $C$ implicit and simply refer to it as the Grothendieck construction.
  Alternatively, this construction can be understood as a coend
  $\IMode{C\Grothendieck{}{\Psi}}\cong \IMode{\int^{c \in C}\Coslice{c}{C}
    \otimes \Discrete{\Psi(c)}}$.}
$\Of{\Grothendieck{}{(-)}}{\Sets^{C}\to\Cats}$ on operators, we have a category
of objects
$\Member{\Pair{\Pair{\Upsilon}{a}}{\vartheta}}{\Grothendieck{}{\Operators}}$ for
$\Member{\vartheta}{\Operators\Pair{\Upsilon}{a}}$ and morphisms
$\Hom{\Grothendieck{}{\Operators}}{\Pair{\Pair{\Upsilon}{a}}{\vartheta}}{\Pair{\Pair{\Upsilon'}{a'}}{\vartheta'}}$
for $\Of{\Pair{\varrho}{f}}{\Pair{\Upsilon}{a}\to\Pair{\Upsilon'}{a'}}$ such
that
$\EqMember{\Lift{\varrho}{f}(\vartheta)}{\vartheta'}{\Operators\Pair{\Upsilon'}{a'}}$.
Equivalently, $\Grothendieck{}{\Operators}$ is the pullback of $\Operators$
along the universal $\Sets$-bundle where $\pi_{\Operators}$ is a discrete
Grothendieck opfibration and $\Forget{-}_{\Singleton}$ is the forgetful functor
from pointed sets:

@@

Operators are defined by specifying the fibers of $\pi_{\Operators}$ in which
they reside. For instance, consider the lambda calculus with a single sort,
$\SortExp$; we give its signature by asserting the following about its
operators:
\begin{gather*}
  \IsOperator{\Upsilon}{\lambda}{
    \MkArity{\MkValence{}{\SortExp}{\SortExp}}{\SortExp}
  }\\
  \IsOperator{\Upsilon}{\mathsf{ap}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{}{}{\SortExp}
    }{\SortExp}
  }
\end{gather*}

@@

\begin{gather*}
  \IsOperator{\Upsilon}{\mathsf{decl}}{
    \MkArity{
      \MkValence{}{}{\SortExp},
      \MkValence{\SortExp}{}{\SortExp}
    }{
      \SortExp
    }
  }\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathsf{get}[u]}{
    \MkArity{}{\SortExp}
  }\\
  \IsOperator{\Upsilon,u:\SortExp}{\mathsf{set}[u]}{
    \MkArity{
      \MkValence{}{}{\SortExp}
    }{
      \SortExp
    }
  }
\end{gather*}

@@

Let the judgment $\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{s}$ presuppose
$\IsMetaCtx{\Omega}$, $\IsSymCtx\Upsilon$, $\IsVarCtx\Gamma$ and $\IsSort{s}$,
meaning that $M$ is an abstract binding tree of sort $s$, with metavariables in
$\Omega$, parameters in $\Upsilon$, and variables in $\Gamma$. Let the judgment
$\IsBTm{\Omega}{\Upsilon}{\Gamma}{E}{v}$ presuppose $\IsValence{v}$.  Then, the
syntax of abstract binding trees (abts) is inductively defined in four rules:

@@

Abts are identified up to $\alpha$-equivalence. Let $\FS{M}$
be the collection of symbols free in $M$, and let $\FV{M}$ be the collection of
variables free $\pi_\Operators$ to operators in $M$. %% TODO: actually define these

@@

Variable substitution in abts is defined inductively by a
pair of judgments, $\IsSubst{N}{x}{M}{M'}$ and $\IsSubst{N}{x}{E}{F}$:

@@

\section{Case Study: Wellformed Sequents}

The representation of telescopes and sequents in a logical framework is
notoriously difficult; whilst it is possible to use higher-order abstract
syntax or abts to encode the binding-structure of telescopes
and sequents, the encoding is sufficiently laborious and obscure that it is not
used in practice.

Crary has demonstrated a first-order encoding of contexts in the logical
framework in bijection with actual LF-contexts, which has been successfully
used in large-scale mechanization efforts, including that of Standard ML and
the Edinburgh Logical Framework itself.

%%% TODO: Cite Crary; cite mechaniziation of Standard ML; cite Martens & Crary
%%% TODO: talk about context encodings in Abella

We will approach the problem of encoding telescopes and sequents from the
\emph{refinements} perspective, where a conservative approximation of the
grammar is first given using the abt logical framework, and then the
correctness of a code is expressed separately in a judgment that refines the
existing specification.

Because we have not committed to using the built-in binding machinery to
express the well-scopedness of telescopes and sequents, we are free to use
\emph{symbols} in order to model the variables in the context. This is actually
quite sensible if we are trying to actually faithfully represent the syntax of
telescopes and sequents, rather than replace them with their counterparts on the
meta-level. It is certainly not intended that we should be able to take a
context $\Match{\Gamma}{x:A}$ and apply the substitution
$\IsSubst{M}{x}{\Gamma}{M:A}$, which is nonsensical; on the other hand, if $x$
had been a \emph{symbol}, all we could do is rename it in the context, since
substitution is not defined for symbols.

This insight leads the way to a simple abt signature for the theory of telescopes
and sequents.
%
\newcommand\SortCtx{\mathsf{ctx}}
\newcommand\SortJdg{\mathsf{jdg}}
\newcommand\SortType{\mathsf{type}}
\newcommand\SortExpr{\mathsf{exp}}
\newcommand\OpNil{\mathsf{nil}}
\newcommand\OpSnoc[1]{\mathsf{snoc}[#1]}
\newcommand\OpVar[1]{\mathsf{var}[#1]}
\newcommand\OpSequent{\mathsf{sequent}}
%
\[
  \begin{array}{l}
    \IsSort{\SortCtx}\\
    \IsSort{\SortExpr}\\
    \IsSort{\SortType}\\
    \IsSort{\SortJdg}\\\\
    \IsOperator{\Upsilon, u:\SortExpr}{\OpVar{u}}{
      \MkArity{}{\SortExpr}
    }\\\\
    \IsOperator{\Upsilon}{\OpNil}{
      \MkArity{}{\SortCtx}
    }\\
    \IsOperator{\Upsilon, u:\SortExpr}{\OpSnoc{u}}{
      \MkArity{
        \MkValence{}{}{\SortCtx},
        \MkValence{}{}{\SortType}
      }{\SortCtx}
    }\\\\
    \IsOperator{\Upsilon}{\OpSequent}{
      \MkArity{
        \MkValence{}{}{\SortCtx},
        \MkValence{}{}{\SortType}
      }{\SortJdg}
    }
  \end{array}
\]

\newcommand\OpPi{\mathsf{pi}}
\newcommand\OpSg{\mathsf{sg}}
\newcommand\OpBool{\mathsf{bool}}
\newcommand\OpSo{\mathsf{isTrue}}

Suppose we have defined encoded a fragment of type theory as well:
\[
  \begin{array}{l}
    \IsOperator{\Upsilon}{\top}{
      \MkArity{}{\SortType}
    }\\
    \IsOperator{\Upsilon}{\bot}{
      \MkArity{}{\SortType}
    }\\
    \IsOperator{\Upsilon}{\OpBool}{
      \MkArity{}{\SortType}
    }\\
    \IsOperator{\Upsilon}{\OpSo}{
      \MkArity{
        \MkValence{}{}{\SortExp}
      }{\SortType}
    }\\
    \IsOperator{\Upsilon}{\OpPi}{
      \MkArity{
        \MkValence{}{}{\SortType},
        \MkValence{}{\SortExp}{\SortType}
      }{\SortType}
    }\\
    \IsOperator{\Upsilon}{\OpSg}{
      \MkArity{
        \MkValence{}{}{\SortType},
        \MkValence{}{\SortExp}{\SortType}
      }{\SortType}
    }\\
  \end{array}
\]

Terms written using the abstract syntax will be difficult to read, so let us
define some notation:
\begin{align*}
  \IMode{\diamond} &\triangleq \OMode{\OpNil}\\
  \IMode{\Gamma, u:P} &\triangleq \OMode{
    \App{\OpSnoc{u}}{\Gamma, P}
  }\\
  \IMode{\Gamma\gg P} &\triangleq \OMode{
    \App{\OpSequent}{
      \Gamma, P
    }
  }\\
  \IMode{`u} &\triangleq \OMode{\OpVar{u}}
\end{align*}

Now, we have the following well-formed sequent:
\[
  \IsAbt{\cdot}{
    u:\SortExp,
    v:\SortExp
  }{\cdot}{
    \diamond,u:\OpBool,v:\App{\OpSo}{`u}%
    \gg%
    \App{\OpSo}{`u}
  }{\SortJdg}
\]

%% TODO: add \nabla operator to close the context
%% TODO: implement refinement judgment for wellscopedness
